---
title: "Exporting Graphs"
execute:
  warning: false
  message: false
---

How do you get graphs out of R in the right size and resolutation? 

## `ggsave()`: the essentials

```{r}
#| eval: false
library(ggplot2)

p <- ggplot(mpg, aes(displ, hwy)) + geom_point() + theme_minimal()

# Vector: for papers (scales without quality loss)
ggsave("figure1.pdf", plot = p, width = 16, height = 10, units = "cm")

# Pixel: for the web and slides
ggsave("figure1.png", plot = p, width = 16, height = 10, units = "cm", dpi = 300)
```

Three parameters matter:

- **`width` / `height`** — the *physical* size. This is what controls how large text appears relative to the plot. Always set these explicitly; the default is "whatever your plotting window happened to be", which is not reproducible.
- **`dpi`** — pixel density, for raster formats only. 300 for print, ~150 for screens.
- **format** — chosen by the file extension. `.pdf` and `.svg` are vector; `.png` and `.jpg` are pixel. Never export a chart as `.jpg`: JPEG compression is designed for photos and puts artifacts around sharp lines and text.

## The most common mistake: fixing size by scaling

If your exported plot has tiny unreadable text, the wrong fix is exporting bigger and scaling down in Word, which shrinks the text further. The right fix: keep the *physical* size close to how it will be printed (a journal text column is ~8–16 cm wide) and increase `base_size` in the theme:

```{r}
#| eval: false
p + theme_minimal(base_size = 14)   # bigger text, same physical size
```

Rule of thumb: text in a figure should end up roughly the same size as the caption text below it.

## Which format for which destination?

| Destination | Format | Settings |
|---|---|---|
| Journal submission | PDF (or EPS if required) | exact column width in cm |
| Word document | PNG | 300 dpi, final physical size |
| Website / blog | PNG or SVG | ~150 dpi, or SVG for sharpness |
| Slides | PNG | 150–200 dpi, 16:9-friendly size |
| Poster (professional print) | PDF | vector; check CMYK requirements |
| Twitter/Bluesky/LinkedIn | PNG | 1200×675 px works well |

## Better rendering: `ragg` and `svglite`

The default PNG device varies across operating systems (different fonts, different anti-aliasing). The [`ragg`](https://ragg.r-lib.org/) package renders identically everywhere and handles custom fonts properly:

```{r}
#| eval: false
library(ragg)
ggsave("figure1.png", plot = p, device = agg_png,
       width = 16, height = 10, units = "cm", dpi = 300)
```

For SVG output use `svglite::svglite` — SVGs stay sharp at any zoom level and are small for line-based graphics, making them ideal for the web.

In a Quarto project you can set this once in `_quarto.yml` instead of per chunk:

```yaml
knitr:
  opts_chunk:
    dev: "ragg_png"
    dpi: 300
    fig-width: 7
    fig-height: 4.5
```

## Fonts in exported figures

If you use a custom font (see the Fonts page), PDF export may silently substitute it. Two solutions: use `ragg` devices (they find system fonts automatically), or embed fonts in PDFs with `cairo_pdf` as the device. Always check the exported file and not the RStudio preview before submitting anywhere.

```{r}
cairo_pdf(
  filename = "figure.pdf",
  width = 8,
  height = 5.5,
  family = "Arial",
  onefile = TRUE
)
```
## Reproducibility checklist

1. Export with code (`ggsave()`), never with the RStudio Export button. Clicks are not reproducible.
2. Fix `width`, `height`, and `dpi` explicitly.
3. Keep an R script or Quarto document that regenerates every figure in the paper from raw data.
4. Name files after the figure they are (`fig1_coefplot.pdf`), and regenerate them all before each submission.
