---
title: "Applied Examples"
---

First, load the packages required for the examples. 

```{r, output=F, message=F}
library(ggplot2)
library(dplyr)
library(broom)
library(margins)
library(ggeffects)
library(patchwork)
library(viridis)

#devtools::install_github("vdeminstitute/vdemdata")
library(vdemdata)
```

# 📊 Introduction to `ggplot2`

`ggplot2` is one of the most powerful and flexible packages in R for creating data visualizations. It follows the **Grammar of Graphics**, meaning it builds plots in layers.

We’ll use the built-in `mpg` dataset for the basic examples.

------------------------------------------------------------------------

## 1. Basic Plot Structure

Every `ggplot2` plot starts with the `ggplot()` function, where you specify:

-   The data frame
-   The aesthetic mappings: `aes()`

```{r}
library(ggplot2)

ggplot(data = mpg, aes(x = displ, y = hwy))
```

This creates a plot object, but it won’t render a chart until you add a layer.

## 2. Add Geometries

You must add a geometry layer (`geom_*()`) to display the data.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

Common geometries:

-   `geom_point()` – scatterplot
-   `geom_line()` – line plot
-   `geom_col()` – bar chart (with values)
-   `geom_bar()` – bar chart (with counts)
-   `geom_boxplot()` – boxplot
-   `geom_histogram()` – histogram

------------------------------------------------------------------------

## 3. Aesthetic Mappings

You can map variables to color, size, shape, or linetype inside `aes()`.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

------------------------------------------------------------------------

## 4. Labels and Titles

Use `labs()` to customize the title, subtitle, and axis labels.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency by Engine Size",
    subtitle = "Colored by Vehicle Class",
    x = "Displacement (L)",
    y = "Highway MPG",
    color = "Class"
  )
```

------------------------------------------------------------------------

## 5. Facets (Small Multiples)

Use `facet_wrap()` to split the data into multiple panels by category.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

------------------------------------------------------------------------

## 6. Themes

Themes control the look of the plot: grid, font, background.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  theme_minimal()
```

Popular themes:

-   `theme_minimal()`
-   `theme_classic()`
-   `theme_light()`
-   `theme_bw()`
-   `theme_void()`

------------------------------------------------------------------------

## 7. Scales

Use `scale_*()` functions to adjust axes, colors, or sizes.

```{r}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  scale_color_viridis_d()
```

------------------------------------------------------------------------

## 8. Full Example

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Fuel Efficiency vs. Engine Size",
    subtitle = "Vehicle class shown by color",
    x = "Engine Displacement (L)",
    y = "Highway MPG",
    color = "Class"
  ) +
  theme_minimal() +
  scale_color_viridis_d()
```

------------------------------------------------------------------------

# Applied Examples for Statistical Models

For the following examples, we use the Varities of Democracy [`(V-Dem)`](https://v-dem.net) dataset.

**Outcomes:**

-   Continuous: `e_gdppc` GDP per Capita

-   Binary: `democracy_binary` (recoded from `v2x_regime`)

**Predictors:**

-   `v2x_jucon` (judicial constraints)

-   `v2x_corr` (control of corruption)

-   `v2x_libdem` Liberal Democracy Index

-   `e_regionpol_6C` (region)

**Get Data:**

```{r}
vdem <- vdemdata::vdem
vdem <- vdem %>%
  mutate(democracy_binary = ifelse(v2x_regime >= 2, 1, 0))
```

## OLS Model

What is the effect of liberal democracy on GDP?

```{r}

ols <- lm(e_gdppc ~ v2x_libdem+v2x_corr+ as.factor(e_regionpol_6C), data = vdem)

broom::tidy(ols, conf.int = TRUE) %>%
  mutate(term = recode(term,
    "v2x_libdem" = "Liberal Democracy",
    "v2x_corr" = "Corruption Control",
    "as.factor(e_regionpol_6C)2" = "Latin America & Caribbean",
    "as.factor(e_regionpol_6C)3" = "Middle East & North Africa",
    "as.factor(e_regionpol_6C)4" = "Sub-Saharan Africa",
    "as.factor(e_regionpol_6C)5" = "Western Europe & North America",
    "as.factor(e_regionpol_6C)6" = "Asia & Pacific"
    # Region 1 is the reference category, not shown in output
  ))%>%
  mutate(term = factor(term, levels = rev(c(
    "Liberal Democracy",
    "Corruption Control",
    "Latin America & Caribbean",
    "Middle East & North Africa",
    "Sub-Saharan Africa",
    "Western Europe & North America",
        "Asia & Pacific",
        "(Intercept)"
  )))) %>%
  ggplot(aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  labs(
    title = "OLS Coefficient Plot: GDP",
    x = "Estimate", y = NULL
  ) +
  theme_minimal()

```

- In `ggplot2`, the order of categorical variables on axes follows the factor level order.
- `recode()` changes the labels; `factor(..., levels = ...)` sets the plotting order.



### Predicted Values (OLS)

```{r}

freexp_preds <- predict_response(ols, terms = c("v2x_libdem"))


ggplot(freexp_preds, aes(x = x, y = predicted)) +
  geom_line()+  
  labs(
    title = "Predicted GDP by Democracy",
    x = NULL,
    y = "Predicted Index Value"
  ) +
  theme_minimal(base_size = 13)

```

-   Computes predicted values based on a regression model you’ve specified.

-   Predictions follow the functional form you impose (e.g., linear relationship, additive terms).


Based on estimated coefficients, it calculates:

-   Fitted/predicted values

-   Confidence intervals (e.g., 95% CI)

### Non-Parametric Smoothing (geom_smooth())

```{r}
vdem_clean <- vdem %>%
  filter(!is.na(v2x_libdem), !is.na(e_gdppc))

ggplot(vdem_clean, aes(x = v2x_libdem, y =e_gdppc)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.3) +
  labs(
    title = "Raw Data Smoothing: GDP by Liberal Democracy",
    x = "Liberal Democracy Index",
    y = "Observed GDP per Capita"
  ) +
  theme_minimal(base_size = 13)

```

-   Applies a data-driven local regression (e.g., LOESS) to the actual data points.
-   Does not assume a global functional form (e.g., linearity).
-   Captures non-linear patterns and noise that are present in the raw data.

### Predicted Values by Region (OLS)

```{r}
    
ggplot(freexp_preds, aes(x = as.factor(x), y = predicted)) +
  geom_point()+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  coord_flip() +
scale_x_discrete(labels = c(
    "1" = "Eastern Europe & Central Asia",
    "2" = "Latin America & Caribbean",
    "3" = "Middle East & North Africa",
    "4" = "Sub-Saharan Africa",
    "5" = "Western Europe & North America",
    "6" = "Asia & Pacific"
  )) +   scale_y_continuous(limits = c(0, 60)) +
  labs(
    title = "Predicted GDP by Region",
    subtitle ="Mean Democracy Index = 0.24",
    x = NULL,
    y = "Predicted Index Value"
  ) +
  theme_minimal(base_size = 13)

```

### Split by Region

```{r}

freexp_preds <- predict_response(ols, terms = c("v2x_libdem", "e_regionpol_6C"))

region_labels <- c(
  "1" = "Eastern Europe & Central Asia",
  "2" = "Latin America & Caribbean",
  "3" = "Middle East & North Africa",
  "4" = "Sub-Saharan Africa",
  "5" = "Western Europe & North America",
  "6" = "Asia & Pacific"
)


ggplot(freexp_preds, aes(x = x, y = predicted, color=group)) +
  geom_line(linewidth=1.1)+  
  labs(
    title = "Predicted GDP by Democracy",
    x = "Democracy Index",
    y = "Predicted GDP Value"
  ) +  scale_color_viridis_d(labels = region_labels, name = "Region") +
  theme_minimal(base_size = 13)

```

## Logit Model: Binary Democracy Outcome

What is the effect of Judicial Constraints on Executive on Democracy?

```{r}
logit_model <- glm(democracy_binary ~ v2x_jucon + v2x_corr  + as.factor(e_regionpol_6C),
                   data = vdem, family = binomial)


broom::tidy(logit_model, conf.int = TRUE)%>%
  mutate(term = recode(term,
    "v2x_jucon" = "Judicial Constraints",
    "v2x_corr" = "Corruption Control",
    "as.factor(e_regionpol_6C)2" = "Latin America & Caribbean",
    "as.factor(e_regionpol_6C)3" = "Middle East & North Africa",
    "as.factor(e_regionpol_6C)4" = "Sub-Saharan Africa",
    "as.factor(e_regionpol_6C)5" = "Western Europe & North America",
    "as.factor(e_regionpol_6C)6" = "Asia & Pacific"
    # Region 1 is the reference category, not shown in output
  )) %>%
  mutate(term = factor(term, levels = rev(c(
    "Judicial Constraints",
    "Corruption Control",
    "Latin America & Caribbean",
    "Middle East & North Africa",
    "Sub-Saharan Africa",
    "Western Europe & North America",
    "Asia & Pacific",
    "(Intercept)"
  )))) %>%
  ggplot(aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  labs(
    title = "Logit Coefficient Plot: Democracy (Binary)",
    x = "Estimate (Log Odds)", y = NULL
  ) +
  theme_minimal()
```

```{r}
dem_preds <- ggpredict(logit_model, terms = c("v2x_jucon [all]", "e_regionpol_6C"))


# Plot using ggplot and viridis color scale
ggplot(dem_preds, aes(x = x, y = predicted, color = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15, color = NA) +
  scale_color_viridis_d(labels = region_labels, name = "Region", end = 0.9) +
  scale_fill_viridis_d(labels = region_labels, name = "Region", end = 0.9) +
  labs(
    title = "Predicted Probability of Democracy by Region",
    x = "Judicial Constraints on Executive (1=No Constraints)",
    y = "Pr(Democracy)"
  ) +
  theme_minimal(base_size = 13)

```

## Interaction

```{r}
interact <- lm(e_gdppc  ~ v2x_corr * democracy_binary,
                      data = vdem)

corr_region_pred <- ggpredict(interact,
                              terms = c("v2x_corr", "democracy_binary"))

# Relabel group variable (0/1) to "Autocracy"/"Democracy"
corr_region_pred$group <- factor(corr_region_pred$group,
  levels = c("0", "1"),
  labels = c("Autocracy", "Democracy")
)

ggplot(corr_region_pred, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  scale_color_viridis_d(name = "Regime Type", end = 0.8) +
  scale_fill_viridis_d(name = "Regime Type", end = 0.8) +
  labs(
    title = "Interaction: Corruption × Regime Type on GDP per Capita",
        subtitle = "0 = less corrupt, 1 = more corrupt",
    x = "Corruption",
    y = "Predicted GDP per Capita"
  ) +
  theme_minimal(base_size = 13)

```

# Marginal Effects

For more details and help interpreting regression models in more depth, I highly recommend the [`marginaleffects`](https://marginaleffects.com) package.

It allows you to:

-   Compute marginal effects, slopes, and predicted probabilities
-   Easily visualize effects, including interactions and uncertainty
-   Supports GLMs, mixed models, multinomials, and more
-   Handles robust standard errors, grouping, and subsets

# Spatial Data

Spatial data in R comes in two main forms:

### 1. **Vector Data**

Represents features as **points, lines, or polygons**.\
Used for things like cities or geocoded events (points), roads (lines), or countries (polygons).

-   **Object class:** `sf` (Simple Features) — modern and tidy-friendly\
-   **Package:** `{sf}`\
-   🛠️ Replaces older `sp` and integrates well with `dplyr`, `ggplot2`, and `tidyverse`

### 2. **Raster Data**

Stores data in **grids/cells**, such as satellite imagery, elevation, or temperature.

-   **Object class:** `SpatRaster` (in {terra}) or RasterLayer (older {raster})
-   **Package:** {terra} (modern), {raster} (legacy)


Extract world map

```{r, message=F}
library(sf)
library(rnaturalearth)
library(mapview)
library(osmdata)


# Load country geometries
world <- ne_countries(scale = "medium", returnclass = "sf")
```



ggplot vs. base plot 

::::: columns
::: {.column width="50%"}


```{r, message=F}
# Plot base map
ggplot(world) +
  geom_sf() +
  labs(title = "World Countries Map") +
  theme_minimal()
```


:::

::: {.column width="50%"}


```{r, message=F}
plot(world$geometry)
```

:::
:::::

 Map a Continuous Variable (GDP per Capita)

```{r, message=F}

# Use Gapminder data (filter for latest year)
gap2007 <- gapminder::gapminder %>%
  filter(year == 2007)

# Join with spatial data
world_gap <- world %>%
  left_join(gap2007, by = c("name" = "country"))

# Plot map
ggplot(world_gap) +
  geom_sf(aes(fill = gdpPercap)) +
  scale_fill_viridis_c(option = "C", na.value = "grey80") +
  labs(title = "GDP per Capita (2007)", fill = "GDP per Capita") +
  theme_minimal()
```

Map a Categorical Variable (Continent)

```{r}
ggplot(world) +
  geom_sf(aes(fill = continent)) +
  labs(title = "Continent Classification", fill = "Continent") +
    scale_fill_viridis_d(option = "D") +
  theme_minimal()
```

Extract a country

```{r}
germany <- world[world$name == "Germany", ]
plot(st_geometry(germany))

```

### Interactive viewing of spatial objects in R

Map population estimates per country

```{r}
mapview(world, zcol ="pop_est")
```

### Mapping of geocoded data

Map geocoded event locations on U.S. Map

```{r, message=F}
### 🧪 Simulate and Map Protest Events Inside U.S. State Borders
us <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "United States of America")


# Simulate random points inside bounding box
set.seed(123)
n_events <- 1000
raw_points <- tibble(
  long = runif(n_events, min = -125, max = -66),
  lat = runif(n_events, min = 25, max = 49),
  protest_mass = sample(50:5000, n_events, replace = TRUE)
)

# Convert to sf points and intersect with US shape
points_sf <- st_as_sf(raw_points, coords = c("long", "lat"), crs = 4326)

points_us <- points_sf[st_intersects(points_sf, us, sparse = FALSE), ]



# Plot result
ggplot() +
  geom_sf(data = us, fill = "grey90", color = "white") +
  geom_sf(data = points_us, aes(size = protest_mass), color = "red", alpha = 0.6) +
  scale_size(range = c(0.25, 3), name = "Protesters") +
  labs(
    title = "Simulated Protest Events in the U.S.",
    x = NULL, y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(25, 50)) +
  theme_minimal()

```

## Overlay on OpenStreetMaps (OSM)

In this example, we scrape prison locations from OSM and map den on an interactive map.

First, get and prepare the data.

```{r, eval=F}
# bounding box of where we want to query data
q <- opq(bbox = st_bbox(st_transform(germany, 4326)))
ger <- opq(bbox = "Germany")

# Build the query of location of prisons
#osmq <- add_osm_feature(q, key = "amenity", value = "prison")
osmq <- add_osm_feature(ger, key = "amenity", value = "prison")

# And then query the data
pris.osm <- osmdata_sf(osmq)

# Make unique points / polygons
pris.osm <- unique_osmdata(pris.osm)

# Get points and polygons 
pris.points <- pris.osm$osm_points
# filter only points within Germany
pris.points <- pris.osm$osm_points |> 
  st_as_sf() |> 
  st_filter(germany)
# Get polygons 
poly <- pris.osm$osm_polygons |> 
  st_as_sf() |> 
  st_filter(germany)
multipoly <- pris.osm$osm_multipolygons |> 
  st_as_sf() |>   st_make_valid() |>
  st_filter(germany)
# Harmonize columns (keep only common ones, or add missing if needed)
common_cols <- intersect(names(pris.points), names(poly)) |> 
  intersect(names(multipoly))

pris.points <- pris.points[, common_cols]
poly <- poly[, common_cols]
multipoly <- multipoly[, common_cols]

# Convert polygons and multipolygons to centroids
poly_pts <- st_centroid(poly)
multipoly_pts <- st_centroid(multipoly)
# Combine all as point geometries
pris_all <- bind_rows(pris.points, poly_pts, multipoly_pts)

```
```{r, echo=FALSE}
# I saved the previous chunk as RDS for faster rendering
#saveRDS(pris_all, "data/prisons.rds")

pris_all <- readRDS("data/prisons.rds")
```

Map the data. Specify the indicator with `zcol = " "`

```{r}
mapview(pris_all, zcol = "name", legend = FALSE)
```


# Nighttime lights

[`BlackMarbleR`](https://worldbank.github.io/blackmarbler/) provides a simple way to use nighttime lights data from NASA’s Black Marble. Black Marble is a NASA Earth Science Data Systems (ESDS) project that provides a product suite of daily, monthly and yearly global nighttime lights.

Prior to using it, you need to register an account [here](https://ladsweb.modaps.eosdis.nasa.gov/) and generate a token.

```{r,output=F, message=F, echo=F}

bearer <- "eyJ0eXAiOiJKV1QiLCJvcmlnaW4iOiJFYXJ0aGRhdGEgTG9naW4iLCJzaWciOiJlZGxqd3RwdWJrZXlfb3BzIiwiYWxnIjoiUlMyNTYifQ.eyJ0eXBlIjoiVXNlciIsInVpZCI6Im1lc2F1dCIsImV4cCI6MTc1Nzg2NTUwOCwiaWF0IjoxNzUyNjgxNTA4LCJpc3MiOiJodHRwczovL3Vycy5lYXJ0aGRhdGEubmFzYS5nb3YiLCJpZGVudGl0eV9wcm92aWRlciI6ImVkbF9vcHMiLCJhY3IiOiJlZGwiLCJhc3N1cmFuY2VfbGV2ZWwiOjN9.C8FLYh6xktgtmWw8Vt15IdYWhs8ew2jF8ei3JbCP8TJX81hLJ9EQHwMWF7upBvwZzvHflga5TnxUsNpBV4owJiMH5MS8MFLSmX_fb7hu7LQfqsyrgjAaMsrQARsD8XVh-bQqYn_YLYSYMIyHpaCcuH0djH2dt491x_aSOAM9wlI0JY8dTm30LrMg0ZBUeR1qh1whmQDK2ZgVgwLXSonRT6Bvb6Lw3NQHVg9rfytUacej6TQV1iRGOaoV3Qdo9KFfcULu0_W3k3T7ULwEqtl4WLL2l8bYd1ZlTtQ6yECG1yIDAWKeSZcX_yjEmotWvaI1AflNP2PTO3hF31D4GgNYug"

```

We are interested in Nighttime Light in Germany. For that, we first need to load a shapefile with the countryborder.

```{r, message=F}
library(blackmarbler)
library(geodata)
library(terra)
library(tidyterra)
library(lubridate)

#### Define NASA bearer token
#bearer <- "BEARER-TOKEN-HERE"
# Get token via account on https://ladsweb.modaps.eosdis.nasa.gov/ 

roi_sf <- gadm(country = "DEU", level=1, path = tempdir()) 

```

```{r, echo=FALSE}
r <- readRDS("data/r.rds")
```

Second, we load the raster data and clip it to the country border. Choose your own date.

```{r, message=F, warning=F}
#### Make raster
#r <- bm_raster(roi_sf = roi_sf,
          #    product_id = "VNP46A3",
            #  date = "2021-10-01",
            #  bearer = bearer, check_all_tiles_exist = FALSE)


#### Prep data
r <- r |> terra::mask(roi_sf)

## Distribution is skewed, so log
r[] <- log(r[] + 1)

##### Map
ggplot() +
  geom_spatraster(data = r) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  labs(title = "Germany Nighttime Lights: October 2021") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  legend.position = "none")
```

### Trends over time

We can also load yearly or monthly data and show e.g. the trend in different states.

```{r, echo=FALSE}
ntl_df <- readRDS("data/germany_ntl.rds")
```

```{r, message=F,  warning=F}
#ntl_df <- bm_extract(roi_sf = roi_sf,
                 #    product_id = "VNP46A4",
                     #date = 2019:2022,
                     #bearer = bearer, check_all_tiles_exist = FALSE)


ntl_df |>
  ggplot() +
  geom_col(aes(x = date,
  y = ntl_mean),
  fill = "darkorange") +
  facet_wrap(~NAME_1) +
  labs(x = NULL,
       y = "NTL Luminosity",
       title = "Germany Admin Level 1: Annual Average Nighttime Lights") +
  scale_x_continuous(labels = seq(2019, 2022, 2),
                     breaks = seq(2019, 2022, 2)) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

### Resources for Spatial Data

-   [Global administrative boundaries (GADM)](https://www.gadm.org/)

-   [Simple Features (SF) Package Description](https://r-spatial.github.io/sf/)

-   [Humanitarian Data Exchange](https://data.humdata.org)

-   [Spatial Data Science](https://r-spatial.org/book/)

-   [BlackmarbleR](https://worldbank.github.io/blackmarbler/)
