---
title: "Advanced Graphics Exercises"
---

```{r, echo=F, output=F}

library(gapminder)
library(ggplot2)
library(dplyr)
library(vdemdata)
library(viridis)
library(countrycode)
library(patchwork)
library(rnaturalearth)

```

# ✍️  Exercises

The following exercises help you explore real-world development data using `ggplot2`, `gapminder` and `V-Dem`.

------------------------------------------------------------------------

## 🧠 Challenge Exercise: Explore a Development Pattern

Before seeing any example plots, try to visualize the following:

> **How does a country's population size relate to life expectancy and income level?**

**Your task**: Create a visualization using the `gapminder` dataset for the year **2007** that captures the relationship between: - `lifeExp`: Life expectancy - `gdpPercap`: GDP per capita - `pop`: Population

You may use any type of plot or mapping of variables (e.g., x-axis, color, size), but **aim to show all three variables clearly**.

⬇️ Fill in the code below. Afterward, we'll compare approaches and interpretations.

```{r}
# ✏️ Add here your own code!

```

##  Exercise 1: Create two Histograms of GDP per Capita 

Use the 2007 data to create two **histogram** that shows how `gdpPercap` and logged `gdpPercap` is distributed across all countries.
Arrange the plots side by side. 


**Hint**: Use `ggplot() + geom_histogram()` and consider `scale_x_log10()`.

⬇️ Fill in your code below:

```{r}
# ✏️ Add here your own code!

```


------------------------------------------------------------------------



## Exercise 2: Faceted Trend Comparison

**Goal:** Compare how GDP per capita and life expectancy evolved over time in three world regions.

**Task:**

- Filter the Gapminder dataset to include only the years 1977, 1987, 1997, and 2007.
- Create a scatterplot of **GDP per capita vs life expectancy**.
- Use **bubble size for population**, **color for continent**, and **facet by year**.

```{r}
# Student solution space here


```

---

##  Exercise 3: Democracy and Freedom of Assocation (V-Dem)

**Goal:** Explore the relationship between democracy and perceived corruption.

**Task:**

- Use V-Dem variables:
  - `v2x_frassoc_thick` (freedom of assocation)
  - `v2x_libdem` (liberal democracy index)
- Visualize their relationship using a **smooth line** or **scatterplot**, and split the plot by **region** using facets.

```{r}
# Student solution space here

```

---

## Exercise 4: Visualizing Interaction Effects

**Goal:** Create an interaction plot for the effect of Legislative Constraints and regime type on Freedom of Expression (V-Dem).

**Task:**

- Build a plot using `ggplot()` with `geom_line()` or `geom_point()`.
- Show the interaction of `v2x_freexp` and a binary democracy variable on GDP (`v2xlg_legcon`).
- Use color or linetype to distinguish regime type.

```{r}
# Student solution space here

```

---

## Exercise 4: Mapping Democracy

**Goal:** Map the latest available democracy index globally using a choropleth map.

**Task:**
- Use the latest year in the V-Dem data.
- Use a spatial join or merge with a world shapefile (e.g., from `rnaturalearth`).
- Plot a map using `geom_sf()` and fill countries by `v2x_libdem`.
- Use `scale_fill_viridis_c()` for color.

```{r}
# Student solution space here


```

---


