---
title: "Advanced Graphics Exercises"
---

```{r, echo=F, output=F}

library(gapminder)
library(ggplot2)
library(dplyr)
library(vdemdata)
library(viridis)
library(countrycode)
library(patchwork)
library(rnaturalearth)

```

# ✍️  Exercises

The following exercises help you explore real-world development data using `ggplot2`, `gapminder` and `V-Dem`.

------------------------------------------------------------------------

## 🧠 Challenge Exercise: Explore a Development Pattern

Before seeing any example plots, try to visualize the following:

> **How does a country's population size relate to life expectancy and income level?**

**Your task**: Create a visualization using the `gapminder` dataset for the year **2007** that captures the relationship between: - `lifeExp`: Life expectancy - `gdpPercap`: GDP per capita - `pop`: Population

You may use any type of plot or mapping of variables (e.g., x-axis, color, size), but **aim to show all three variables clearly**.

⬇️ Fill in the code below. Afterward, we'll compare approaches and interpretations.

```{r}
# ✏️ Add here your own code!
library(viridis)  

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(
    x = gdpPercap,
    y = lifeExp,
    size = pop,
    color = continent
  )) +
  geom_point(alpha = 0.8) +
  scale_x_log10() +
  scale_size(range = c(2, 20), guide = "none") +
  scale_color_viridis_d(option = "D", end=0.95) +
  labs(
    title = "Life Expectancy vs GDP per Capita (2007)",
    subtitle = "Bubble size = population, color = continent",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy",
    color = "Continent"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 15)
  )

```

##  Exercise 1: Create two Histograms of GDP per Capita 

Use the 2007 data to create two **histogram** that shows how `gdpPercap` and logged `gdpPercap` is distributed across all countries.
Arrange the plots side by side. 


**Hint**: Use `ggplot() + geom_histogram()` and consider `scale_x_log10()`.

⬇️ Fill in your code below:

```{r}
# ✏️ Add here your own code!

gap2007 <- gapminder %>% filter(year == 2007)

# A: Raw histogram (no transformation)
hist_raw <- ggplot(gap2007, aes(x = gdpPercap)) +
  geom_histogram(binwidth = 2000, fill = "steelblue", color = "white") +
  labs(
    title = "A. GDP per Capita (Raw Scale)",
    x = "GDP per Capita",
    y = "Count"
  ) +
  theme_classic(base_size = 13)


# B: Histogram with log-scaled x-axis
hist_logscale <- ggplot(gap2007, aes(x = gdpPercap)) +
  geom_histogram(binwidth = 0.1, fill = "darkred", color = "white") +
  scale_x_log10() +
  labs(
    title = "B. GDP per Capita (Log10 Scale)",
    x = "GDP per Capita (log10)",
    y = "Count"
  ) +
  theme_classic(base_size = 13)

# Combine both using patchwork
hist_raw + hist_logscale + plot_annotation(
  title = "Distribution of GDP per Capita in 2007",
  tag_levels = 'A'
)

```


------------------------------------------------------------------------



## Exercise 2: Faceted Trend Comparison

**Goal:** Compare how GDP per capita and life expectancy evolved over time in three world regions.

**Task:**

- Filter the Gapminder dataset to include only the years 1977, 1987, 1997, and 2007.
- Create a scatterplot of **GDP per capita vs life expectancy**.
- Use **bubble size for population**, **color for continent**, and **facet by year**.

```{r}
# Student solution space here

gap_filtered <- gapminder %>% 
  filter(year %in% c(1977, 1987, 1997, 2007))

ggplot(gap_filtered, aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  facet_wrap(~ year) +
  scale_color_viridis_d() +
  scale_size(range = c(2, 20), guide = "none") +
  labs(
    title = "GDP vs Life Expectancy Over Time",
    x = "GDP per Capita (log)",
    y = "Life Expectancy",
    size = "",
    color = "Continent"
  ) +
  theme_minimal()

```

---

##  Exercise 3: Democracy and Freedom of Assocation (V-Dem)

**Goal:** Explore the relationship between democracy and perceived corruption.

**Task:**

- Use V-Dem variables:
  - `v2x_frassoc_thick` (freedom of assocation)
  - `v2x_libdem` (liberal democracy index)
- Visualize their relationship using a **smooth line** or **scatterplot**, and split the plot by **region** using facets.

```{r}
# Student solution space here

vdem_filtered <- vdem %>%
  filter(!is.na(v2x_frassoc_thick), !is.na(v2x_libdem), !is.na(e_regionpol_6C)) %>%
  mutate(region = factor(e_regionpol_6C))

# Custom region labels
region_labels <- c(
  "1" = "Western Europe & North America",
  "2" = "Eastern Europe & Central Asia",
  "3" = "Latin America & Caribbean",
  "4" = "Sub-Saharan Africa",
  "5" = "Middle East & North Africa",
  "6" = "Asia & Pacific"
)


ggplot(vdem_filtered, aes(x = v2x_frassoc_thick, y = v2x_libdem)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "darkred") +
  facet_wrap(~ region, labeller = as_labeller(region_labels)) +
  labs(
    title = "Relationship between Corruption and Liberal Democracy",
    x = "Freedom of Assocation (v2x_frassoc_thick)",
    y = "Liberal Democracy Index (v2x_libdem)"
  ) +
  theme_minimal()
```

---

## Exercise 4: Visualizing Interaction Effects

**Goal:** Create an interaction plot for the effect of Legislative Constraints and regime type on Freedom of Expression (V-Dem).

**Task:**

- Build a plot using `ggplot()` with `geom_line()` or `geom_point()`.
- Show the interaction of `v2x_freexp` and a binary democracy variable on GDP (`v2xlg_legcon`).
- Use color or linetype to distinguish regime type.

```{r}
# Student solution space here

vdem_interact_data <- vdem %>%
  filter(
    !is.na(v2xlg_legcon),
    !is.na(v2x_freexp),
    !is.na(v2x_regime)
  ) %>%
  mutate(
    regime_type = ifelse(v2x_regime == 1, "Democracy", "Autocracy")
  )

ggplot(vdem_interact_data, aes(x = v2xlg_legcon, y = v2x_freexp, color = regime_type)) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_viridis_d(option = "D") +
  labs(
    title = "Interaction: Legislative Constraints and Regime Type",
    subtitle = "Do stronger legislative constraints improve freedom of expression?",
    x = "Legislative Constraints on Executive (v2xlg_legcon)",
    y = "Freedom of Expression Index (v2x_freexp)",
    color = "Regime Type"
  ) +
  theme_minimal(base_size = 13)

```

---

## Exercise 4: Mapping Democracy

**Goal:** Map the latest available democracy index globally using a choropleth map.

**Task:**
- Use the latest year in the V-Dem data.
- Use a spatial join or merge with a world shapefile (e.g., from `rnaturalearth`).
- Plot a map using `geom_sf()` and fill countries by `v2x_libdem`.
- Use `scale_fill_viridis_c()` for color.

```{r}
# Student solution space here

latest_vdem <- vdem %>%
  filter(year == max(year), !is.na(v2x_libdem)) %>%
  mutate(iso3c = countrycode(country_name, "country.name", "iso3c"))

world <- ne_countries(scale = "medium", returnclass = "sf")

world_map <- left_join(world, latest_vdem, by = c("iso_a3" = "iso3c"))

ggplot(world_map) +
  geom_sf(aes(fill = v2x_libdem)) +
  scale_fill_viridis_c(option = "D", na.value = "grey90") +
  labs(
    title = "Liberal Democracy Index (Most Recent Year)",
    fill = "v2x_libdem"
  ) +
  theme_minimal()

```

---


